Parsing Postfix log files

### 2007-04-03 doc/logparser.tex
Major work: two thirds of the full algorithm description is written; it definitely requires editing and proof reading by someone not intimately connected with the code. There are supporting changes to the flow chart; they are minor but pervasive. Added assumptions section to Background. Explanatory comment added to the flowchart to clarify what it describes. Added actions in detail section; the skeleton is there, but it needs to be filled out. Expanded the limitations and improvements sections. Various punctuation corrections based on reading Eats, Shoots and Leaves. Various minor improvements

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible.  This paper describes the process required, documenting
    the parsing algorithm and rules, and explaining the difficulties
    encountered, with reference to an implementation which stores data
    gathered from the logs in an SQL database for later processing.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be selectively applied, combined and
bypassed on a per-client, per-recipient or per-sender basis, allowing
varying levels of defense and or permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions, with the ultimate in flexibility being the
facility to consult an external process which can implement whatever logic
is required: some users can be restricted to sending mail on the third
Tuesday after pay day only, if that is of any use.\footnote{This example
may actually be useful in a payroll system.  More commonly encountered
scenarios are restricting posting to large mailing lists (e.g. sending
weekly offers to customers) to out of office hours, checking SPF records
(see http://www.openspf.org/ or
http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details), or
rate limiting clients.}  Administrators can also supply their own rejection
messages, to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

@@

@@

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, detailed later in Section
        \ref{actions-in-detail}.} but there is a facility for the
        specifying the order of the rules, so that more specific rules can
        be tried first.

@@

    \item [regex] The regex to match the log line against.  The regex will
        first have several keywords expanded: this simplifies reading and
        writing rules, avoids needless repetition of complex regex
        components, and allows the components to be corrected and/or
        improved in one location.  For efficiency the keywords are expanded
        and every rule's regex is compiled before attempting to parse the
        log file - otherwise each regex would be recompiled each time it
        was used, resulting in an 468.95\% slowdown.

@@

    \item [result\_data, connection\_data] Sometimes rules need to fake a
        piece of data which isn't present in the log line: e.g. setting
        smtp\_code when mail is accepted.  The format is the same as
        result\_cols and connection\_cols, except that arbitrary data can
        be specified rather than just a number.

@@

        A full list can be found in Section \ref{actions-in-detail}.

@@

This example rule matches the message Postfix logs when it rejects mail
from a sender address where the domain has neither an MX record nor an A
record, i.e. mail could not be delivered to the sender's address.  
For full details see
http://www.postfix.org/postconf.5.html\#reject\_unknown\_sender\_domain

@@

While the rules are more than double (treble with 2.3.x rules?) the size of
the algorithm, the rules are quite simple, and each rule is completely
independent of its fellows.  The parser is significantly more complicated,
and highly internally interdependent.

@@

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox (or program; there is very little
        difference between the two): sometimes they are aliased and
        need to be delivered to an address (or addresses) on another
        server.  When this occurs a child mail will be injected into the
        postfix queue, but without the explicit logging smtpd or sendmail
        injected mails have, so the source is not immediately discernible;
        from a strictly linear reading of the logs it \textit{usually}
        appears as if the child mail has appeared from thin air.
        Subsequently the parent mail will log the creation of the child
        mail:

@@

\subsection{Algorithm flow chart}

\includegraphics{logparser-flow-chart.ps}
\label{flow-chart}

@@

@@

@@

    \item The program should save state at the end of each run, and reload
        it at the start of the next run, to properly cope with mails which
        span logs.  This might also require purging old mails once they've
        been in the state tables for too long.

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10, this may possibly have changed in Postfix 2.3, or the
        upcoming 2.4.}

@@

>>>>

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible.  This paper describes the process required, documenting
    the parsing algorithm and rules, explaining the difficulties
    encountered, with reference to an implementation which stores data
    gathered from the logs in an SQL database for later processing.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be selectively applied, combined and
bypassed on a per-client, per-recipient or per-sender basis, allowing
varying levels of defense and or permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions, with the ultimate in flexibility being the
facility to consult an external process which can implement whatever logic
is required: some users can be restricted to sending mail on the third
Tuesday after pay day only, if that is of any use.\footnote{This example
may actually be useful in a payroll system.  More commonly encountered
scenarios are restricting posting to large mailing lists (e.g.\ sending
weekly offers to customers) to out of office hours, rate limiting clients,
or checking SPF records (see http://www.openspf.org/ or
http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details)}
Administrators can also supply their own rejection messages, to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use \texttt{grep(1)} to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
the journey is simple, but the other 10\% of the time requires 90\% of the
code.\footnote{These numbers don't have a solid scientific basis, they're
based on gut feeling from writing and debugging the software.}

@@

\subsection{Assumptions}

The algorithm described and the program implementing it make a small number
of (hopefully safe and reasonable) assumptions:

    \item The logs are whole and complete; nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging unable to cope with the volume of logs).

    \item Postfix logs sufficient information to make it possible to
        accurately reconstruct the actions it has taken.

    \item The Postfix queue has not been tampered with, causing unexplained
        appearance or disappearance of mail.

In some ways this task is similar to reverse engineering a program, or
replicating a black box program based solely on its inputs and outputs.
Although the source code is available, reading and understanding it would
require a significant investment of time:

\begin{tabular}[]{lll}

    Postfix 2.3.8   & Postfix 2.4.0 &                   \\
    82224           & 83965         & lines of code     \\
    67146           & 68675         & lines of comments \\
    17647           & 18069         & lines are blank   \\
    167017          & 170709        & lines in total    \\

@@

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, details are provided later
        in Section~\ref{actions-in-detail}.} but there is a facility for
        the specifying the order of the rules, so that more specific rules
        can be tried first.

@@

    \item [regex] The regex to match the log line against.  The regex will
        first have several keywords expanded: this simplifies reading and
        writing rules, avoids needless repetition of complex regex
        components, and allows the components to be corrected and/or
        improved in one location.  For efficiency the keywords are expanded
        and every rule's regex is compiled before attempting to parse the
        log file --- otherwise each regex would be recompiled each time it
        was used, resulting in an 468.95\% slowdown.

@@

    \item [result\_data, connection\_data] Sometimes rules need to fake a
        piece of data which isn't present in the log line: e.g.\ setting
        smtp\_code when mail is accepted.  The format is the same as
        result\_cols and connection\_cols, except that arbitrary data can
        be specified rather than just a number.

@@

        A full list can be found in Section~\ref{actions-in-detail}.

@@

This example rule matches the message Postfix logs when it rejects mail
from a sender address where the domain has neither an MX record nor an A
record, i.e.\ mail could not be delivered to the sender's address.  For full
details see
http://www.postfix.org/postconf.5.html\#reject\_unknown\_sender\_domain

XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
PERHAPS WHEN THE ALGORITHM IS EXPLAINED\@? XXX

@@

While the rules are more than double (treble with 2.3.x rules?) the size of
the algorithm, the rules are quite simple and each rule is completely
independent of its fellows.  The parser is significantly more complicated,
and highly internally interdependent.

@@

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox (or program; there is very little
        difference between the two): sometimes they are aliased and
        need to be delivered to an address (or addresses) on another
        server.  When this occurs a child mail will be injected into the
        postfix queue, but without the explicit logging smtpd or sendmail
        injected mails have, so the source is not immediately discernible;
        from a strictly linear reading of the logs it \textit{usually\/}
        appears as if the child mail has appeared from thin air.
        Subsequently the parent mail will log the creation of the child
        mail:

@@

\subsection{Flow chart}

\label{flow-chart}

This flow chart shows the paths the data representing a mail/connection can
take through the parser algorithm.

\includegraphics{logparser-flow-chart.ps}

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
I felt that the best way to implement this was to maintain state
information for every unfinished mail and manipulate it appropriately for
each log line encountered.  The parser thus requires both a method of
mapping log lines to the correct mail and a method of specifying the action
the log line represents.  The former is achieved by using the pid of the
smtpd during the initial phase, then switching to the queueid once the mail
has been accepted.  The latter uses the action field of the rule which
matched the log line, executing the code in the function named by the
action.  From a high level viewpoint this design shares significant
similarity with how an editor maintains state for multiple files and
performs the actions the user initiates on the correct file.

XXX USE A DISPATCH TABLE TO MAKE THIS ACCURATE XXX

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The flow chart in section~\ref{flow-chart} should
be consulted also.

\subsubsection{Mail enters the system}

\label{mail-enters-the-system}

Everything starts off with a mail entering the system, by local submission
via postdrop/sendmail, by SMTP, or by re-injection due to forwarding.
Local submission is the simplest of the three: a queueid is assigned
immediately and the sender address is logged (action: pickup;
flowchart:~2).  SMTP is more complicated:

    \item First there is a connection from the remote client
        (action: connect; flowchart:~1).

    \item This is followed either by rejections (action: save\_by\_pid;
        flowchart:~4) or acceptance of one or more mails (action: clone,
        then save\_by\_pid; flowchart:~5,~4), or a combination of both.

    \item The client disconnects (action: disconnect; flowchart:~6).  As
        soon as a mail is accepted it is assigned a queueid; if there is no
        queueid Postfix didn't accept any mail over that connection, it
        rejected all attempts due to the configured restrictions.  Absence
        of a queueid is checked for during disconnection handling (action:
        still disconnect; flowchart:~7) and if none is found the data is
        cleaned up and entered in the database (action: still disconnect;
        flowchart:~8; subroutines: fixup\_connection and
        commit\_connection), then removed from memory (action: still
        disconnect; flowchart:~9).

    \item If one or more mails were accepted there will be more log entries
        later.

Re-injection due to forwarding sadly lacks explicit log lines of its own;
there are two implicit indications:

    \item The more commonly occurring indicator is qmgr selecting a mail
        with a previously unseen queueid for delivery (action:
        qmgr\_chooses\_mail; flowchart:~3), in which case a new data
        structure will be created.  The mail will be marked as faked; this
        flag should be subsequently cleared.  The data is then saved
        (action: still qmgr\_chooses\_mail; flowchart:~3).

    \item This indicator will be seen for every mail that is re-injected
        but is less common in identifying re-injected mail because it
        typically occurs after the indicator explained above.  Local
        delivery re-injects the mail and logs a successful delivery rather
        than delivering directly to a mailbox or program (action: track;
        flowchart:~11).  In this case the mail may already have been
        created (by qmgr\_chooses\_mail, above); if not a new data
        structure will be created.  In both cases the new mail is marked as
        a child of the parent.

Re-injection is somewhat awkward to explain because it overlaps both the
mail entering and mail delivery sections.  See
section~\ref{tracking-re-injected-mail} for a full discussion.

\subsubsection{Mail delivery}

The obvious counterpart to mail entering the system is mail leaving the
system, whether it is by deletion, bouncing or delivery.  All three are
handled in exactly the same way:

    \item The sender and recipient addresses will be logged separately
        (action: save\_by\_queueid; flowchart:~10).

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~11) --- see
        section~\ref{tracking-re-injected-mail}.

    \item Eventually the mail will be delivered (even if re-injected),
        bounced, or deleted by the administrator (action: commit;
        flowchart:~12).  This is the last line in the logs for this
        particular mail (it may be indirectly referred to if it was
        re-injected).  If it is neither parent nor child of re-injection
        the data is cleaned up and entered in the database (action: still
        commit; flowchart:~14), then deleted from memory (action: still
        commit; flowchart:~15).  For re-injected mails see 
        section~\ref{tracking-re-injected-mail}.

I'd like to reiterate that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is when the mail is
re-injected due to forwarding, and that deserves its own section, which is
coming up next.

\subsubsection{Tracking re-injected mail}

\label{tracking-re-injected-mail}

It's probably apparent by now that tracking re-injected mails is the single
most complex part of the parser.

@@

    \item [ignore] Ignore this line, so that all log lines we're interested
        in are parsed, but this line doesn't have any effect.

    \item [connect] Handle a connection from a remote client.  Complains if
        a connection already exists for this client

    \item [disconnect]

    \item [save\_by\_pid]

    \item [save\_by\_queueid]

    \item [commit]

    \item [track]

    \item [restriction\_start]

    \item [qmgr\_chooses\_mail] 
        however qmgr selects every mail for delivery, so most of the time
        this action merely falls through to saving the information from the
        log line (action: save\_by\_queueid; flowchart:~10).

    \item [pickup]

    \item [clone]

@@

    \item The program should save state at the end of each run, and reload
        it at the start of the next run, to properly cope with mails which
        span logs.  This might also require purging old mails once they've
        been in the state tables for too long, though the presence of very
        old mails indicates a bug in either the implementation or the
        specification.

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix
        2.2.10; this may possibly have changed in Postfix 2.3, or the
        upcoming 2.4.}  It should be reasonably simple to write a policy
        server which causes Postfix to log a warning containing the HELO
        hostname when the DATA command is accepted.

    \item The algorithm does not distinguish between mails where one or
        more mails are rejected and a subsequent mail is accepted; it will
        appear in the database as one mail with lots of rejections followed
        by acceptance.  I don't believe it's possible to make this
        distinction given the data Postfix logs, though it might be
        possible to write a policy server to provide additional logging.

@@

    \item Write the policy server referred to in the third limitation
        above.
